Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-03-07 rmd/manuscript.Rmd
update: manuscript

State transitions shed lights on how FCL changes occurred. In the chain motif, the "additive" mechanism (i.e., the addition of higher trophic levels) was always responsible for the changes in FCL. With disturbance, state transition occurred from "B" to "B + C" (low productivity) or "B + C" to "B + C + P" (high productivity) as ecosystem size or complexity increases. Exceptionally, the state transition of "B + C + P" $\rightarrow$ "B + C" occurred as ecosystem complexity increased with no disturbance and low productivity, a scenario in which ecosystem complexity decreased FCL.

In the weak and strong omnivory motifs, ecosystem size and complexity increased FCL through different mechanisms. Ecosystem size increased FCL primarily through the additive mechanism, facilitating the state transition of "B" $\rightarrow$ "B + P" (low productivity) or "No species" $\rightarrow$ "B + P" (high productivity), particularly with frequent disturbance (Figure XXX). In contrast, the insertion and additive mechanisms were responsible for the effect of ecosystem complexity on FCL, depending on simulation scenarios. With weak omnivory, the transition from "B + P" to "B + C + P" (insertion) was predominant when productivity was low (Figure XXX). However, the additive mechanism ("No species" $\rightarrow$ "B + P") seems to be dominant in the high productivity scenario (Figure XXX). In the strong omnivory motif, transition patterns depended on disturbance frequency. When disturbance was rare ($p_m = 0.05$), the insertion mechanism ("B + P" $\rightarrow$ "B + C + P") drove the FCL change (Figure XXX). As disturbance frequency increased, the primary process switched to the additive mechanism ("B" $\rightarrow$ "B + P" or "No species" $\rightarrow$ "B + P"; Figure XXX).

@@

State transitions associated with increased productivity depended on the motif structure. In the chain motif, the transition from "B + C" to "B + C + P" (i.e., additive) was dominant regrardless of disturbance frequency. The introduction of omnivory allowed the transition between "B + P" and "B + C + P", the primary direction of the transition differed between weak and strong omnivory. With weak omnivory, the transition of "B + C + P" $\rightarrow$ "B + P" was prevalent and shortened FCL. However, this transition was marginal in the strong omnivory motif. Instead, productivity significantly increased the proportion of "No species" state (Figure XXX), especially in small and/or simple ecosystems (Figure XXX).

@@

The most significant finding in this study is the synergistic effects between ecosystem geometry and disturbance. Our no disturbance scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independently of ecosystem size and complexity (except the chain motif with low productivity). However, as the frequency of intense disturbance increases, small and/or simple networks became unable to maintain long food chains. Here, increased ecosystem size and complexity may give greater resistance to disturbance through different mechanisms. Ecosystem size simply increases the "total amount" of disturbance refugia, from which dispersing propagules recolonize impacted habitat patches. In the meantime, ecosystem complexity may increases the connectivity to disturbance refugia. Indeed, branching shortens the mean distance between patches and therefore may enhance immigration from less undisturbed patches. It is also conceivable that the proportion of disturbance refugia increased in more branched networks. However, we did not observe a clear proportional increase in disturbance refugia with increasing branching probability. Therefore, each of geometric features support long food chains through distinct mechanisms. While empirical studies rarely looked at the influence of ecosystem complexity on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled primarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. More empirical research is needed to confirm our theoretical predictions.

The synergy between ecosystem geometry and disturbance has also an important implication for why the dynamical stability hypothesis [@pimmNumberTrophicLevels1977] has been least supported by the literature. Although increased disturbance frequency consistently reduced FCL, its magnitude depended strongly on ecosystem geometry. This observation should provide a possible explanation for the limited support for the dynamical stability hypothesis in field studies. Although FCL research often assumes that the effects of environmental controls are independent of one another, exploring potential interactions among them could provide useful insights into what controls FCL in nature.

@@

The effect of productivity was context-dependent, supporting previous theoretical predictions [@postProximateStructuralMechanisms2007; @kondohFoodchainLengthAdaptive2009; @wardMechanisticTheoryAquatic2017]. While productivity clearly increased FCL in the chain motif, this effect was vague (strong omnivory) or even reversed (weak omnivory) in omnivory motifs. Intraguild predation plays an important role in explaining this context dependent effect. In the high productivity scenario with omnivory, the increased abundance of the basal species enhances the intraguild predator. This may subsequently facilitate the local exclusion of the intraguild prey through increased predation pressure, if the attack rate is sufficiently high [@postProximateStructuralMechanisms2007; @kondohFoodchainLengthAdaptive2009; @wardMechanisticTheoryAquatic2017]. Indeed, in the weak omnivory motif (the high value of the attack rate of P on C), we observed the state transition from "B + C + P" to "B + P" that shortened FCL. However, in the strong omnivory motif, the transition was

Ecosystem size and complexity led to the distinct patterns of state transitions. The state transitions with increased ecosystem size are likely associated with increased immigration. The total number of dispersing propagules increases with increasing ecosystem size, providing opportunities to maintain longer food chains in connected patches. The exception was the strong omnivory motif with rare disturbance, in which the insertion mechanism operated. This is likely because rare disturbance inhibited the exclusion of the intraguild prey by suppressing the abundance of the intraguild predator. Similarly, increased immigration may be generally responsible for the state transitions with increasing ecosystem complexity. Habitat patches are closer in more branched networks as branching compresses the network configuration. However, unlike ecosystem size, ecosystem complexity does not increase the total number of dispersing propagules - rather, it enhances connectivity among habitat patches. This difference may explain the distinct patterns of state transitions along the gradients of ecosystem size and complexity.

>>>>

State transitions shed lights on how FCL changes occurred. In the chain motif, the "additive" mechanism (i.e., the addition of higher trophic species) was always responsible for the changes in FCL. With disturbance, state transition occurred from "B" to "B + C" (low productivity) or "B + C" to "B + C + P" (high productivity) as ecosystem size or complexity increases. Exceptionally, the state transition of "B + C + P" $\rightarrow$ "B + C" occurred as ecosystem complexity increased with no disturbance and low productivity, a scenario in which ecosystem complexity decreased FCL.

In the weak and strong omnivory motifs, ecosystem size and complexity increased FCL through different mechanisms. Ecosystem size increased FCL primarily through the additive mechanism, facilitating the state transition of "B" $\rightarrow$ "B + P" (low productivity) or "No species" $\rightarrow$ "B + P" (high productivity), particularly with frequent disturbance (Figure XXX). In contrast, the insertion (i.e., the insertion of intermediate trophic species) and additive mechanisms were responsible for the effect of ecosystem complexity on FCL, depending on simulation scenarios. With weak omnivory, the transition from "B + P" to "B + C + P" (insertion) was predominant when productivity was low (Figure XXX). However, the additive mechanism ("No species" $\rightarrow$ "B + P") seems to be dominant in the high productivity scenario (Figure XXX). In the strong omnivory motif, transition patterns depended on disturbance frequency. When disturbance was rare ($p_m = 0.05$), the insertion mechanism ("B + P" $\rightarrow$ "B + C + P") drove the FCL change (Figure XXX). As disturbance frequency increased, the primary process switched to the additive mechanism ("B" $\rightarrow$ "B + P" or "No species" $\rightarrow$ "B + P"; Figure XXX).

@@

State transitions associated with increased productivity depended on the motif structure. In the chain motif, the transition from "B + C" to "B + C + P" (i.e., additive) was dominant regardless of disturbance frequency. The introduction of omnivory allowed the transition between "B + P" and "B + C + P", the primary direction of the transition differed between weak and strong omnivory. With weak omnivory, the transition of "B + C + P" $\rightarrow$ "B + P" was prevalent and shortened FCL regardless of disturbance frequency (Figure XXX). However, this transition was marginal in the strong omnivory motif. Instead, productivity significantly increased the proportion of "No species" state (Figure XXX) in disturbed scenarios, especially in small and/or simple ecosystems (Figure XXX).

@@

The most significant finding in this study is the synergistic effects between ecosystem geometry and disturbance. Our no disturbance scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independently of ecosystem size and complexity (except the chain motif with low productivity). However, as the frequency of intense disturbance increases, small and/or simple networks became unable to maintain long food chains. Here, increased ecosystem size and complexity may give greater resistance to disturbance through different mechanisms. Ecosystem size simply increases the "total amount" of disturbance refugia, from which dispersing propagules recolonize impacted habitat patches. In the meantime, ecosystem complexity may increase the connectivity to disturbance refugia. Indeed, branching shortens the mean distance between patches and therefore may enhance immigration from less disturbed patches. It is possible that the proportion of disturbance refugia increased in more branched networks. However, we did not observe a clear proportional increase in disturbance refugia with increasing branching probability. Collectively, each of geometric features likely support long food chains through distinct mechanisms, which may also explain the different patterns of state transitions (additive vs. insertion mechanisms) [@postProximateStructuralMechanisms2007].

While empirical studies rarely looked at the influence of ecosystem complexity on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled primarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. More empirical research is needed to confirm our theoretical predictions.

The synergy between ecosystem geometry and disturbance has also an important implication for why the dynamical stability hypothesis has been least supported by the literature [@takimotoEnvironmentalDeterminantsFoodchain2013]. Although increased disturbance frequency consistently reduced FCL, its magnitude depended strongly on ecosystem geometry. This observation should provide a possible explanation for the limited support for the dynamical stability hypothesis in field studies. Although FCL research often assumes that the effects of environmental controls are independent, exploring potential interactions among them could provide useful insights into environmental controls of FCL in nature.

@@

The effect of productivity was motif-dependent, supporting previous theoretical predictions [@postProximateStructuralMechanisms2007; @kondohFoodchainLengthAdaptive2009; @wardMechanisticTheoryAquatic2017]. While productivity clearly increased FCL in the chain motif through an additive mechanism, the productivity effect was vague (strong omnivory) or even reversed (weak omnivory) in omnivory motifs. Intraguild predation plays an important role in explaining this context dependency. In the high productivity scenario with omnivory, the increased abundance of the basal species enhances the intraguild predator. This may subsequently facilitate the local exclusion of the intraguild prey through increased predation pressure, if the attack rate is sufficiently high [@postProximateStructuralMechanisms2007; @kondohFoodchainLengthAdaptive2009; @wardMechanisticTheoryAquatic2017]. Indeed, in the weak omnivory motif (the high value of the attack rate of P on C), we observed the state transition from "B + C + P" to "B + P" with shortened FCL regardless of disturbance frequency. However, this was not the case in the strong omnivory motif; the increased proportion of "No species" state had a strong impact on the minimum FCL, and this transition was mediated by disturbance. This result suggests that, although the strong omnivory motif is innately stable with high productivity (see one-patch simulations; Figure XXX), external disturbance undermines the sensitive balance of predator-prey interactions with the abrupt increase of "No species" state, which must be countered by immigration from less disturbed habitats to maintain food chains.

It is therefore probable that different processes underlie the productivity-driven enforcement of the FCL-geometry relationships among food web motifs. In the chain and strong omnivory motifs, increased productivity magnified the disproportionate impact of disturbance on small and/or simple ecosystems because they seem less stable with limited supply of immigrants and external perturbations. With the weak omnivory motif, productivity facilitated the spatial spread of the intraguild predator in small ecosystems, where dispersing propagules can reach wider areas of a network with local extirpation of the intraguild prey.
